---
title: "Predicting Customer Churn"
subtitle: "DSCI 575"
author: "Jayden Bradbury, Jonah Fallon"
date: "`r Sys.Date()`"
format: pdf
---

```{r}
#| echo: false
#| include: false
#| label: front-matter
#| warning: false
# clean up & set default chunk options
rm(list = ls())
knitr::opts_chunk$set(echo = FALSE)

# packages
library(tidyverse)
library(caret)
library(ggplot2)
library(corrplot)
library(dplyr)
library(DiagrammeR)
library(caret)
library(xtable)
library(knitr)
library(ggthemes)
library(kableExtra)


# read in data 

train <- read.csv("playground-series-s6e3/train.csv")
test  <- read.csv("playground-series-s6e3/test.csv")


```

# 1. Project Description

Customer churn refers to when a customer stops doing business with a company. High churn rates reduce revenue, increase acquisition costs, and negatively impact long term growth. The goal of this project is to predict the probability that a customer will churn based on customer data. By identifying high risk customers early, businesses can implement targeted retention strategies. This can reduce revenue loss and support data-driven decision-making. This can be especially helpful for new businesses where keeping customers is crucial in the early stages of a business.

In this project, we analyze a company's customer base and the data from each customer to predict whether or not that customer will churn. We also calculate the probability that a customer will churn given certain variables. 

## 1.1 Research Question

Using the data provided from the "Predict Customer Churn" Playground [1], we answer the following research question:

**What is the probability that a customer will churn given the customer's data?**

## 1.2 Data and Variables

As stated above, the data we are using is from the "Predict Customer Churn Playground from Kaggle [1]. In the downloaded package, we obtain a training dataset and a testing dataset with the variables in @tbl-variables and @tbl-variables2.

@tbl-variables and @tbl-variables2 show the variables used in this project and in our analysis:

<!--
table below will include the variable names, type, and a short description of that variable
-->

```{r}
#| label: tbl-variables
#| echo: false
#| tbl-cap: "Summary of Variables Used in Analysis"
variable.desc <- data.frame(Name = c("ID",
                                     "Gender",
                                     "SeniorCitizen",
                                     "Partner",
                                     "Dependents",
                                     "Tenure",
                                     "PhoneService",
                                     "MultipleLines",
                                     "InternetService",
                                     "OnlineSecurity",
                                     "OnlineBackup",
                                     "DeviceProtection",
                                     "TechSupport",
                                     "StreamingTV",
                                     "StreamingMovies"))
variable.desc$Type <- c("Numerical",
                        "Binary",
                        "Binary",
                        "Binary",
                        "Binary",
                        "Numerical",
                        "Binary",
                        "Categorical",
                        "Categorical",
                        "Categorical",
                        "Categorical",
                        "Categorical",
                        "Categorical",
                        "Categorical",
                        "Categorical")
variable.desc$Notes <- c("A unique identifier assigned to each individual customer in the dataset.",
                         "Indicates whether the customer is female (0) or male (1).",
                         "A binary flag indicating if the customer is younger than 65 (0) or at least 65 (1).",
                         "Indicates whether the customer is married or has a domestic partner (1) or not (0).",
                         "Specifies if the customer lives with children, parents, or other family members who rely on them financially (1) or not (0)",
                         "The total number of months the customer has been with the company",
                         "Indicates whether the customer has a home phone service subscription (1) or not (0).",
                         "Specifies if the customer has multiple phone lines, a single line, or no phone service",
                         "The type of internet connection the customer uses (DSL, Fiber Optic, or none).",
                         "Indicates if the customer has an active subscription for online security software.",
                         "Specifies if the customer uses the company's online data backup service.",
                         "Indicates if the customer has insurance or a protection plan for their hardware.",
                         "Specifies if the customer has access to premium technical support services.",
                         "Indicates whether the customer uses their internet service to stream television programs (if they have internet service).",
                         "Indicates whether the customer uses their internet service to stream movies (if they have internet service)."
                         )
knitr::kable(variable.desc, format = "latex", booktabs = TRUE,
             col.names = c("Name", "Type", "Notes"),
             longtable = FALSE) %>%
  kable_styling( full_width = FALSE) %>%
  column_spec(3, width = "8cm") %>% 
  add_header_above(c(" " = 3))
```

\newpage

```{r}
#| label: tbl-variables2
#| echo: false
#| tbl-cap: "Summary of Variables Used in Analysis (cont.)"
variable.desc <- data.frame(Name = c("Contract",
                                     "PaperlessBilling",
                                     "PaymentMethod",
                                     "MonthlyCharges",
                                     "TotalCharges",
                                     "Churn"))
variable.desc$Type <- c("Categorical",
                        "Binary",
                        "Categorical",
                        "Numerical",
                        "Numerical",
                        "Binary (Target)")
variable.desc$Notes <- c("The length of the customer's current service agreement (Month-to-month, one year, or two year).",
                         "Indicates if the customer has opted to receive their bills electronically (1) rather than by mail (0).",
                         "The primary method the customer uses to pay their monthly bill (Mailed check, credit card (automatic), electronic check, bank transfer (automatic)).",
                         "The specific amount billed to the customer on a monthly basis.",
                         "The cumulative amount the customer has been charged over the entire duration of their tenure.",
                         "The target variable indicating whether the customer stopped using the company's services (only in the 'training' dataset)."
                         )
knitr::kable(variable.desc, format = "latex", booktabs = TRUE,
             col.names = c("Name", "Type", "Notes"),
             longtable = FALSE) %>%
  kable_styling( full_width = FALSE) %>%
  column_spec(3, width = "8cm") %>% 
  add_header_above(c(" " = 3))
```

# 2. Exploratory Data Analysis (EDA)

<!--
idk how, but we need to try and look at some eda, even though our XGBoost model includes every variable. this can be where we mention how we scaled the variables and why, and our correlation plot.
--> 

# 3. Statistical Analysis 

<!--
our model, reasoning for picking our model, and the model output/summary
-->

# 4. Conclusion & Results

<!--
where we answer the research question formally. need to provide a probability
-->


# 5. Additional Considerations

<!--
ethical considerations, biases, etc.
-->


# 6. References

<!--
reference list
-->

[1]“Predict Customer Churn,” @kaggle, 2026. https://www.kaggle.com/competitions/playground-series-s6e3 (accessed Apr. 28, 2026)‌ 

# Technical Appendix 

<!--
code important to our analysis, but that we shouldn't include in the main body of the report. This will be like validation and other models we tried.
-->

\newpage

### R Script

<!--
this is where all of our code will be outputted, because we don't want code chunks in the main body of our report.
-->

```{r ref.label=c('front-matter', 'tbl-variables', 'tbl-variables2')}
#| echo: true
#| eval: false

# Reprinted code chunks used previously for analysis
```
