---
title: "Predicting Customer Churn"
subtitle: "DSCI 575"
author: "Jayden Bradbury, Jonah Fallon"
date: "`r Sys.Date()`"
format: pdf
---

```{r}
#| echo: false
#| include: false
#| label: front-matter
#| warning: false
# clean up & set default chunk options
rm(list = ls())
knitr::opts_chunk$set(echo = FALSE)

# packages
library(tidyverse)
library(caret)
library(ggplot2)
library(corrplot)
library(dplyr)
library(DiagrammeR)
library(caret)


# read in data 

train <- read.csv("playground-series-s6e3/train.csv")
test  <- read.csv("playground-series-s6e3/test.csv")


```

# 1. Project Description

Customer churn refers to when a customer stops doing business with a company. High churn rates reduce revenue, increase acquisition costs, and negatively impact long term growth. The goal of this project is to predict the probability that a customer will churn based on customer data. By identifying high risk customers early, businesses can implement targeted retention strategies. This can reduce revenue loss and support data-driven decision-making. This can be especially helpful for new businesses where keeping customers is crucial in the early stages of a business.

In this project, we analyze 


## 1.1 Research Question

**What is the probability that a customer will churn given the customer's data?**

## 1.2 Variables


<!--
table below will include the variable names, type, and a short description of that variable
-->

```{r}
#| label: tbl-variables
#| echo: false
#| tbl-cap: "Summary of Variables Used in Analysis"
variable.desc <- data.frame(Name = c())
variable.desc$Type <- c()
variable.desc$Notes <- c()
knitr::kable(variable.desc, format = "latex", booktabs = TRUE,
             col.names = c("Name", "Type", "Notes"),
             longtable = FALSE) %>%
  kable_styling( full_width = FALSE) %>%
  column_spec(3, width = "8cm") %>% 
  add_header_above(c(" " = 3))
```

# 2. Exploratory Data Analysis (EDA)

<!--
idk how, but we need to try and look at some eda, even though our XGBoost model includes every variable. this can be where we mention how we scaled the variables and why, and our correlation plot.
--> 

# 3. Statistical Analysis 

<!--
our model, reasoning for picking our model, and the model output/summary
-->

# 4. Conclusion & Results

<!--
where we answer the research question formally. need to provide a probability
-->

# 5. References

<!--
reference list
-->

# 6. Additional Considerations

<!--
ethical considerations, biases, etc.
-->

# Technical Appendix 

<!--
code important to our analysis, but that we shouldn't include in the main body of the report. This will be like validation and other models we tried.
-->

\newpage

### R Script

<!--
this is where all of our code will be outputted, because we don't want code chunks in the main body of our report.
-->

```{r ref.label=c('front-matter')}
#| echo: true
#| eval: false

# Reprinted code chunks used previously for analysis
```
